The form of uncertainty affects selection for social learning

### 2022-06-03 ms.tex
Updates from Overleaf

\title{Uncertainty in social learning evolution}

@@

Social learning is essential to human and other species' everyday life and survival.
Social learning enhances problem solving when acquiring information from others is
more efficient than learning on one's own~\cite{Laland2004}. Theory predicts that
social learning should be favored in uncertain
environments~\cite{BoydRicherson1985,Henrich1998}, up to a
point~\cite{Rogers1988,Feldman1996}; this prediction has received some empirical
support across species~\cite{McElreath2005,Kendal2018,Allen2019}.  However, the
meaning of the term \emph{uncertainty} often conflates environmental variability,
spatial heterogeneity, ambiguity in behavioral payoff structure, and other
variability that could plausibly be interpreted as \emph{uncertainty}.  Furthermore,
many models of social learning evolution do not account for individual-level
cognition~\cite{Heyes2016}, though humans clearly have evolved cognitive mechanisms
for dealing with uncertainty generally~\cite{Vul2014,Gershman2019,Schulz2019}.
Social learning evolution theory tends to be organized around an increasing number
of modeling choices~\cite[Figure 1]{Kendal2018} instead of around questions of when
and why social learning is adaptive, in which contexts.  Therefore, we developed
an evolutionary, empirically-motivated, agent-based model that operationalized
four principal uncertainty components to generate predictions of when 
social learning would evolve in which circumstances, and why. Answering this
question is critical for understanding human behavior adapting to a rapidly changing
world, both to mitigate existential threats~\cite{Moya2020,Jones2021}, and to
captialize on new opportunities such as transitioning to clean energy use and
production~\cite{NatureEnergyEditorialPromisesPremises2018,Brisbois2022}.

\mt{This paragraph is a first try to introduce the four uncertainty dimensions
with examples to illustrate the problem we are answering.
I outlined the next paragraph to get into more details of social learning:
how cultural evolution works; conformity, success-biased, etc. Cristina, please
edit this.} 
To gain a more systematic understanding of uncertainty and social learning
evoluiton, we identified four common classes of ecological
uncertainty (``principal components'' of uncertainty) used as factors
in social learning evolution models: (1) environmental variability; (2) payoff
ambiguity (difference between best behavior choice and others); (3) 
selection set size (number of ecologically possible behaviors); and 
(4) the lifespan of agents. These were chosen because modeling and empirical 
study designs across taxa tend use one or more of these uncertainty dimensions.
Environmental variability is sometimes modeled as catastrophic and 
happens only once in an individual's lifetime~\cite{Rogers1988}, or may be
modeled as occurring with a fixed or variable
probability~\cite{Feldman1996,McElreath2005}---in these examples environments
can only be in one of two states. Payoff ambiguity sometimes takes the form
of differences in expected payoffs arising from different
behaviors or strategies~\cite{Enquist2007,Rendell2010}
or by varying the standard deviation of payoffs~\cite{McElreath2005}. 
Selection set size often takes only one value: 
for example, in empirical studies of \emph{bombus terresteris} (bumble
bee)~\cite{Baracchi2018} and \emph{parus major} (great tit)~\cite{Aplin2017},
experimenters provided two behaviors for
the bees or birds to perform, respectively, each yielding greater or smaller
payoffs depending on experimental treatment. Human studies have used two
or three possible behaviors~\cite{McElreath2005,Toyokawa2019}.
Modeling studies have used two 
behaviors~\cite{Feldman1996,Rendell2010} and four behaviors~\cite{Enquist2007}. 
The age or lifespan of individuals was
found to affect on social learning in both \emph{parus
major}~\cite{Aplin2017} and \emph{poecilia reticulata} 
(guppies)~\cite{Leris2016} \mt{Need better connection between lifespan and 
age, or better references to support use of lifespan}.
Shorter lifespans mean individuals
die more uncertain of which behaviors are optimal, and pass that uncertainty on
to their offspring. To answer our research questions about the combined effect
of various forms of uncertainty on social learning evolution, we developed an
agent-based model that incorporates all four of
these uncertainty variables. We systematically vary these parameters
to understand and predict their effects on social learning evolution.

@@

\paragraph{Meaning of uncertainty in ecology and cognitive and social sciences } (JAMIE)
We need to support our claim that these four classes of uncertainty really 
count as forms of uncertainty and show how it is consistent across ecology,
and maybe connect to uncertainty in cognitive and social sciences more broadly.

@@

\begin{figure}
  \caption{Agents are initialized as randomly social learners or not, with their
  payoff observations all initialized to zero (A). Then agents begin selecting
and performing behaviors and accumulating payoffs, which goes on for $L$
timesteps (B). After $L$ time steps, agents are selected to reproduce,
social learner children learn from a member of their parent's generation, and
the previous generation dies off (C). The simulation stops if children are all
social or asocial learners (i.e.\ the system reaches fixation), 
otherwise repeat another generation and evolution (3).}
  \label{fig:schematic}
  \centering
    \includegraphics[width=\textwidth]{Figures/IntraInterGenerationalDynamics.pdf}
\end{figure}

@@

Several patterns emerged in changes to the shape of social learning suppression over
environmental variability, $u$, across different uncertainty contexts defined by the
payoff ambiguity, selection set size, and effective lifespan
(Figure~\ref{fig:mainResults}). When payoff ambiguity and selection set size were
both small ($\pilow = 0.1$ and $B=2$) social learning only evolved when $u$ is small
since individual learning is relatively efficient in this case
(Figure~\ref{fig:mainResults}, upper left). As $\pilow$ increased the suppression of
$\meansl$ flattened over $u$, which was more pronounced for smaller $L$ (rows of
Figure~\ref{fig:mainResults}).  When $L$ is smaller, agents have fewer opportunities
to learn through trial and error, and so it is more difficult for them to discern
which strategy, social or asocial, is optimal. Increased selection set size led to
both a shift in the critical value of $u$ when $\meansl$ begins to decrease, and
flattened social learning suppression. This is because possibly misleading 
social information is less consequential due
to (1) larger $B$ is a more difficult problem for trial-and-error learning and
(2) a misidentified optimal behavior is weighted less relative to smaller
$B$, meaning agents are more likely to explore behaviors identified as non-optimal
when $B$ is larger (columns of Figure~\ref{fig:mainResults}).  
Increased $B$ led to more pronounced flattening of social learning suppression when
$L$ was small, apparently again because of the limited data available to evolution,
making individual outcomes more path dependent.

>>>>

\title{The form of uncertainty affects selection for social learning}

@@

\cm{Uncertainty -> social learning widespread belief. But variously operationalized or not real uncertainty. Selling points of this paper 1) real uncertainty, 2) 4 kinds uncertainty, 3) cognition (soft-max temperature varies how much explores).}

Social learning enhances problem solving when acquiring information from others is more efficient than learning on one's own~\cite{Laland2004}. However, social learning can also lead individuals astray if, for example, they are copying outdated information. Theoretical models have helped clarify the circumstances under which social learning should be evolutionarily favored. For example, environments that change across time favor learning from others (as opposed to having a genetically fixed behavior) ~\cite{BoydRicherson1985,Henrich1998}, so long as they are not so variable that the social information reflects outdated and maladaptive behavior

point~\cite{Rogers1988,Feldman1996}. \cm{check that the citations still accurate}. This theoretical work has motivated much empirical research and helped predict behavior in several species ~\cite{McElreath2005,Kendal2018,Allen2019}.

From this literature, the idea that \emph{uncertainty} supports social learning has proliferated, but the term often conflates several features of the environment, including temporal variability,
spatial heterogeneity, ambiguity in behavioral payoff structure, and difficulty of the task.  In translating mathematical models to verbal predictions, the formal meaning of terms like \emph{uncertainty} is often lost, facilitating such conflation. Furthermore, as the number of formal models of social learning have expanded, an ever increasing number
of modeling choices ~\cite[Figure 1]{Kendal2018} and formalizations of \emph{uncertainty} has made it difficult to compare across models and consolidate our understanding of the contexts in which why social learning is adaptive.

To address these concerns we developed an empirically-motivated, evolutionary agent-based model that operationalizes four principal components of uncertainty. We use this integrated model to generate predictions about the kinds of uncertainty that favor social learning and to clarify the logic of why this would be the case. Answering this question is critical for understanding human behavior adapting to a rapidly changing world, both to mitigate existential threats~\cite{Moya2020,Jones2021}, and to capitalize on new opportunities such as transitioning to clean energy use and production~\cite{NatureEnergyEditorialPromisesPremises2018,Brisbois2022}.

 \cm{I moved this for now: \emph{"Furthermore, many models of social learning evolution do not account for individual-level
cognition~\cite{Heyes2016}, though humans clearly have evolved cognitive mechanisms
for dealing with uncertainty generally~\cite{Gershman2019,Schulz2019}."} because I think it interrupted the flow and didn't seem directly linked to the goal of understanding aspects of uncertainty. However, if this is one of the main features of the model that we want to highlight as an improvement over previous work, then we probably have to say something more about it. Perhaps I just wasn't seeing the connection though, and the more accurate cognitive model DOES add to our ability to study the consequences of uncertainty. If that's the case, the argument should be sharpened. If not, maybe we put this info when the cognitive model is introduced.}\mt{Agreed. I put a first draft of a better version of this paragraph closer to/at the end of the Intro.}

\mt{This paragraph is a first try to introduce the four uncertainty dimensions
with examples to illustrate the problem we are answering.
I outlined the next paragraph to get into more details of social learning:
how cultural evolution works; conformity, success-biased, etc. Cristina, please
edit this.} 
To systematically study features of uncertainty and its consequences for social learning, we identified four common classes of ecological uncertainty (``principal components'' of uncertainty) used in evolutionary models: (1) temporal environmental variability; (2) payoff
ambiguity (difference between best behavior choice and others); (3) 
selection set size (number of ecologically possible behaviors); and 
(4) the lifespan of agents (number of individual learning opportunities).
These were chosen because modeling and empirical 
studies across taxa tend use one or more of these uncertainty dimensions [Figure 1].

\cm{Other choices: softmax: existing models underestimate the power of learning from experience can improve both individual and helps social learners recover from outdated / misleading social info , no conformism: but that's ok because we're giving social learning the best chance of success 
no horizontal transmission, no spatial variation}

Temporal environmental variability is usually modeled as an independent probability of the environment changing its state each generation ~\cite{Feldman1996,McElreath2005}. The probabilities of environmental change are usually fixed ~\cite{BoydRicherson1985, Kendal2009}, but they have also been modeled as deterministic cycles ~\cite{Feldman1996} or as variable rates \cm{who does this?}. Furthermore, the consequences of environmental change can range from mild to  catastrophic. In the latter case a change of environment corresponds to no one having the adaptive behavior for the environment ~\cite{Rogers1988}.

or may be
modeled as occurring with a fixed or variable
probability---in these examples environments
can only be in one of two states. Payoff ambiguity sometimes takes the form
of differences in expected payoffs arising from different
behaviors or strategies~\cite{Enquist2007,Rendell2010}
or by varying the standard deviation of payoffs~\cite{McElreath2005}. 
Selection set size often takes only one value: 
for example, in empirical studies of \emph{bombus terresteris} (bumble
bee)~\cite{Baracchi2018} and \emph{parus major} (great tit)~\cite{Aplin2017},
experimenters provided two behaviors for
the bees or birds to perform, respectively, each yielding greater or smaller
payoffs depending on experimental treatment. Human studies have used two
or three possible behaviors~\cite{McElreath2005,Toyokawa2019}.
Modeling studies have used two 
behaviors~\cite{Feldman1996,Rendell2010} and four behaviors~\cite{Enquist2007}. 
The age or lifespan of individuals was
found to affect on social learning in both \emph{parus
major}~\cite{Aplin2017} and \emph{poecilia reticulata} 
(guppies)~\cite{Leris2016} \mt{Need better connection between lifespan and 
age, or better references to support use of lifespan}.
Shorter lifespans mean individuals
die more uncertain of which behaviors are optimal, and pass that uncertainty on
to their offspring. To answer our research questions about the combined effect
of various forms of uncertainty on social learning evolution, we developed an
agent-based model that incorporates all four of
these uncertainty variables. We systematically vary these parameters
to understand and predict their effects on social learning evolution.

@@

Social learning is the adaptive icing on the cake of intelligent adaptations to ecological uncertainty. To deal with uncertainty as individuals, human experiment participants have been observed to update their tendency for exploring alternative behaviors based on how much uncertainty there is globally. Furthermore, correlations in uncertainty between potential behaviors, humans tend to adopt directed exploration strategies that favor testing behaviors with greater observed payoff variance \cite{Wilson2014,Gershman2019}. Many models of social learning assume little to nothing about the cognitive capabilities of their agents (Cristina unpack this in another paragraph, using the uncertainty studies spreadsheet?). We do not do this because \emph{blackboxing}, or ignoring, cognitive mechanisms is “bad” \cite[p. 658]{Heyes2016, Kendal2018}, but because it is theoretically meaningful. In order to understand how social learning evolves as an adaptation to uncertainty, the model is strengthened by assuming a more biologically plausible, and evolutionarily accurate, model of cognition capable of maximizing payoffs under uncertainty for an individual. In our model we do not impose any structure in randomness between behaviors, so our cognitive model is only sensitive to total randomness. Specifically, we will adopt the softmax distribution for function for selecting a behavior based on its observed payoffs–greater payoff variance means any one behavior is more likely to be selected.

\paragraph{Meaning of uncertainty in ecology (and cognitive and social sciences?) } (JAMIE, PAUL?)
We need to support our claim that these four classes of uncertainty really 
count as forms of uncertainty and show how it is consistent across ecology,
and maybe connect to uncertainty in cognitive and social sciences more broadly.

@@

\begin{figure}
  \caption{Agents are randomly initialized as social learners or not, with their
  payoff observations all initialized to zero (A). Then agents begin selecting
and performing behaviors and accumulating payoffs, which goes on for $L$
timesteps (B). After $L$ time steps, agents are selected to reproduce,
social learner children learn from a member of their parent's generation, and
the previous generation dies off (C). The simulation stops if children are all
social or asocial learners (i.e.\ the system reaches fixation), 
otherwise repeat another generation and evolution (3).}
  \label{fig:schematic}
  \centering
    \includegraphics[width=\textwidth]{Figures/IntraInterGenerationalDynamics.pdf}
\end{figure}

@@

Several patterns emerged in changes to the shape of social learning suppression over
environmental variability, $u$, across different uncertainty contexts defined by the
payoff ambiguity, selection set size, and effective lifespan
(Figure~\ref{fig:mainResults}). When payoff ambiguity and selection set size were
both small ($\pilow = 0.1$ and $B=2$) social learning only evolved when $u$ is small
since individual learning is relatively efficient in this case
(Figure~\ref{fig:mainResults}, upper left). As $\pilow$ increased the suppression of
$\meansl$ flattened over $u$, which was more pronounced for smaller $L$ (rows of
Figure~\ref{fig:mainResults}).  When $L$ is smaller, agents have fewer opportunities
to learn through trial and error, and so it is more difficult for them to discern
which strategy, social or asocial, is optimal. Increased selection set size led to
both a shift in the critical value of $u$ when $\meansl$ begins to decrease, and
flattened social learning suppression. This is because possibly misleading 
social information is less consequential due
to (1) larger $B$ is a more difficult problem for trial-and-error learning, and
(2) a misidentified optimal behavior is weighted less relative to smaller
$B$, meaning agents are more likely to explore behaviors identified as non-optimal
when $B$ is larger (columns of Figure~\ref{fig:mainResults}).  \mt{Support this by calculating the difference between weights when B=2 and B=10 and a social learner learns outdated information}
Increased $B$ led to more pronounced flattening of social learning suppression when
$L$ was small, apparently again because of the limited data available to evolution,
making individual outcomes more path dependent.
